Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario

### 2021-08-29 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton-V0.tex
Edits and file with tracked changes for resubmission to SSM

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Changes in accessibility to emergency and community food services during COVID-19 and implications for low income populations in Hamilton, Ontario},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Changes in accessibility to emergency and community food
services during COVID-19 and implications for low income populations in
Hamilton, Ontario}
    \author[Some University]{Author One}
   \ead{author1@example.com} 
    \author[Another University]{Author Two\corref{1}}
   \ead{author2@example.com} 
    \author[Some University]{Author Three}
   \ead{author3@example.com} 
    \author[Some University]{Author Four}
   \ead{author4@example.com} 
      \address[Some University]{Department, Street, City, State, Zip}
    \address[Another University]{Department, Street, City, State, Zip}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  In this paper we analyze the changes in accessibility to community and
  emergency food services before and during the COVID-19 pandemic in the
  City of Hamilton, Ontario. Many of these food services are the last
  line of support for households facing food insecurity; as such, their
  relevance cannot be ignored in the midst of the economic upheaval
  caused by the pandemic. Our analysis is based on the application of
  balanced floating catchment areas and concentrates on households with
  lower incomes (\textless CAD40,000, approximately the Low Income
  Cutoff Value for a city of Hamilton's size). We find that
  accessibility was low to begin with in suburban and exurban parts of
  the city; furthermore, about 14\% of locations originally available in
  Hamilton closed during the pandemic, further reducing accessibility.
  The impact of closures on the level of service of the remaining
  facilities, and on accessibility, was disproportionate, with
  system-wide losses exceeding 39\%. Those losses were geographically
  and demographically uneven. While every part of the city faced a
  reduction in accessibility, inner suburbs fared worse in terms of loss
  of accessibility. As well, children (age \(\le 18\)) appear to have
  been impacted the most.
  \end{abstract}

@@

The emergence of COVID-19, the worst public health crisis since the 1918
flu pandemic, has exposed important social and economic fault lines, and
pre-existing patterns of inequality appear to have been exacerbated.
Along several other dimensions of stress (e.g., accessibility to health
care facilities, Ghorbanzadeh et al., 2021; Pereira et al., 2021a), this
seems to be the case for food insecurity as well (Laborde et al., 2020).
According to Statistics Canada (2020a), in the early stages of the
pandemic almost 15\% of individuals reported living in a household that
faced food insecurity; the risk of food insecurity was substantially
higher for households with children. The difference between households
with and without children was significant, and 11.7\% of households with
children indicated that ``food didn't last and {[}there was{]} no money
to get more'' sometimes or often, compared to 7.3\% of households
without children; likewise, 13\% of households with children indicated
that they ``{[}c{]}ouldn't afford balanced meals'' sometimes or often,
compared to 8.8\% of households without children. Additionally, Men and
Tarasuk (2021) report that about 25\% of individuals who experienced job
insecurity (a relatively common occurrence during the pandemic) also
experienced food insecurity associated with COVID-related disruptions to
employment, financial hardship, and use of food charity.

The impacts of food insecurity during the pandemic are alarming, since
diet-related diseases, such as obesity, heart-disease, and diabetes,
were already critical public health concerns in Canada prior to COVID-19
(Boucher et al., 2017). While emergency food services are not
necessarily a stable solution to food insecurity and in fact may
encourage a retrenchment of neoliberal policy (Wakefield et al., 2013),
in reality provide a resource of last instance to households in
precarious situations (Bazerghi et al., 2016). As a mid-size city
grappling with deindustrialization, Hamilton exhibits high rates of
poverty and use of emergency food services. As recently as 2019, the
Hamilton Hunger Report (HFS, 2019) noted that food banks in the city
recorded the highest number of visitors in the past 29 years; a rate of
increase greater than population growth. Most troubling, approximately
40\% of all visitors were children.

It is known that urban food environments, within which people make their
daily food choices, are essential in influencing eating behaviours and
health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to emergency and community
food services. Previous work has explored differences in
\emph{accessing} food banks, such as how some households utilize food
banks over short periods of time while others regularly utilize food
banks as longer-term resource (e.g., Enns et al., 2020). In addition,
transportation and locational considerations have been raised as key
issues in food bank accessibility in previous qualitative research
(Smith-Carrier et al., 2017). However, we are not aware of any research
that has focused on estimating or capturing this geographic component of
accessibility.

The study of place-based geographic accessibility is concerned with
capturing the potential to reach destinations of value using the
transportation network (Páez et al., 2012). Indeed, the Government of
Canada's recent Food Policy (Agriculture and Canada, 2019) has made
``access'' to healthy food a priority for Canadian communities and
previous research suggests that such accessibility plays a key role in
user satisfaction with food bank service delivery (Holmes et al., 2018).
However, as with research into the prevalence of food deserts,
accessibility to food banks is unlikely to be evenly distributed, and
variation throughout a city can be expected due to transportation
network characteristics and the spatial distribution of service
locations and the population they are meant to serve. Furthermore,
policy responses to the COVID-19 pandemic likely have added to the
distress of vulnerable households. Non-pharmaceutical interventions
during the pandemic involving restrictions in mobility have increased
the friction of travel, in particular by transit on which low income
populations are more reliant (e.g., DeWeese et al., 2020). At the same
time, the pandemic has created additional stress for the operators of
food banks through disruptions in the supply chain (e.g., McKay et al.,
2021) as well as concerns surrounding the delivery of service in safe
conditions and possible cancellation of food service programs.

For this study, we aim to look at how the landscape of emergency food
and related services (e.g.~low-cost or free meal service providers)
available in Hamilton, Ontario, changed during the pandemic. Did the
number of open services diminish? If so, what was the accessibility to
emergency and community food services before the pandemic from the
perspective of low income households, and how has it changed during the
pandemic with respect to geographic access and congestion at remaining
sites? And finally, who are most likely to have been impacted by changes
in the accessibility landscape? This paper first looks at the
distribution of emergency and community food services before and during
the pandemic. Then, we use the balanced floating catchment area approach
of Paez et al. (2019) to investigate the accessibility situation. For
this, we adopt a fully disaggregated approach based on parcel-level
data. Socio-economic and demographic data are drawn from the latest
Census of Canada (2016), whereas travel information is from the most
recent regional travel survey from 2016. This paper follows reproducible
research recommendations (see Brunsdon and Comber, 2020), and the
research was conducted using open source tools for transportation
analysis (Lovelace, 2021). The data and code needed to reproduce the
analysis are available in a public repository (see Supplemental
Material).

\hypertarget{food-insecurity-and-emergency-and-communal-food-services-in-canada}{%
\section{Food Insecurity and Emergency and Communal Food Services in
Canada}\label{food-insecurity-and-emergency-and-communal-food-services-in-canada}}

Food insecurity is the inability to acquire and consume an adequate
amount or good quality food, leading to inadequate nutrient intake
(Kirkpatrick and Tarasuk, 2008) and poorer physical and mental health
outcomes (Ramsey et al., 2011; Seligman et al., 2010; Stuff et al.,
2004). In this regard, food insecurity is a major population health
concern, particularly among Canadians at socio-economic disadvantage
(Bazerghi et al., 2016). Official government surveys such as the
Household Food Security Survey Module (HFSSM), the Canadian Community
Health Surveys (CCHS), the Longitudinal and International Study of
Adults (LISA), and official classifications determined by Health Canada
in relation to socio-demographic variables offer some insight into food
insecurity in Canada (Gundersen et al., 2018; Kirkpatrick and Tarasuk,
2008; Tarasuk and Vogt, 2009).

Nationally, analysis of the 2011-2012 CCHS has previously revealed that
food insecurity impacts approximately 12.3\% of Canadian households
(Tarasuk et al., 2014). Using the same data, Tarasuk et al. (2019) found
higher odds of food insecurity amongst households relying on social
assistance, those without a university degree or with children under the
age of 18, and individuals that lived alone, renters, and those
identifying as Aboriginal. While surveys revealed that only 20 to 30
percent of those experiencing food insecurity were found to frequent
food banks in Canada (Tarasuk et al., 2014), pre-pandemic research from
Ottawa (Enns et al., 2020) and Vancouver (Black and Seto, 2020) suggests
that long-term users tend to be older, have health or mobility
challenges, live in large households, and are less likely to have
employment income. In terms of geography, previous research conducted at
the provincial scale using data from the 2011-2012 CCHS found that the
prevalence of food insecurity ranged across the country from 11.8\% of
households in Ontario to 41\% of households in Nunavut (Tarasuk et al.,
2019).

Food banks - sometimes also referred to as `food pantries' and `food
shelves' - originated as a community response to aid those with
inadequate food by voluntarily offering them meals and ingredients
(Loopstra and Tarasuk, 2012; Riches, 2002). Although in their origin
food banks were meant to provide a temporary solution to accommodate
those in hunger due to job retrenchments and economic downfalls since
the 1980s, over time many have evolved into a community practice to
secure food supplies for those in need (Loopstra and Tarasuk, 2012;
Wakefield et al., 2013). In Canada, the number of food banks has
steadily increased in the past few decades (Wakefield et al., 2013). The
largest database of food banks and their use comes from the non-profit
association Food Banks Canada (FBC), which conducts an annual assessment
through its affiliated members. FBC's 2018 Hunger Count report (FBC,
2018) (the most recent available) listed 1,830 member food banks across
the country, and found that Canadians visited food banks 1.1 million
times in March of 2018. Of those accessing food banks, certain
population characteristics tend to be over-represented compared to
national totals from the 2016 Canadian Census of Population. According
to FBC's 2018 data, single-adult households represent 45\% of those
utilizing food banks despite making up 28\% of Canada's population, 19\%
are single-parent households (compared to 10\% nationally), and 35\% of
those using food bank services are children aged 0-18 even though their
share of Canada's national population is approximately 20\%. In
addition, 59\% of households accessing food banks list social or
disability assistance as their primary source of income.

The COVID-19 pandemic has led to an increase in the number of households
living in food insecurity in Canada. Survey results from Statistics
Canada from May of 2020 suggest that 14.7\% of the population was living
in food insecurity in the past 30 days, up from 10.5\% in 2017-2018
(Statistics Canada, 2020a). Recent data from FBC (FBC, 2020) showed that
52\% of member food banks reported an increase in usage in March of 2020
when initial lockdown restrictions were put in place across much of the
country. The pandemic also created significant staffing issues with 42\%
of food banks reporting a reduction in volunteers. However, 53\% of food
banks later reported a decrease in use into the summer of 2020 which FBC
members attributed to emergency financial support programs from the
federal government. Nevertheless, some of these benefit programs were
temporary. Although more recent statistics on food bank use in Hamilton
in 2020 and 2021 are not yet available, data from the Daily Bread Food
Bank (DBFB, 2020) in neighbouring Toronto for August 2020 shows visits
climbing 51\% year-over-year in that city, which suggests that many
households in Hamilton are likely to turn to food bank services to meet
their needs.

Beyond traditional conceptualizations of food banks as providers of
emergency food assistance, other community food services also play an
important role in decreasing food insecurity. The scope and objectives
of food banks can vary by region and by country, and these organizations
can include not only prepared meals and aliments for emergency food
supply, but also shared spaces to connect in community gardens and
community kitchens (Wakefield et al., 2013). However, the efficacy of
these programs in reducing food insecurity differs by the type of
service offered. For example, previous qualitative research in the
Toronto region has questioned the capacity of community kitchens to
improve the food security of low-income households due to their limited
scale of operations, and un-subsidized kitchens were found to be
particularly inaccessible to to families living in severe poverty
(Tarasuk and Reynolds, 1999). However, other food access options such as
no-cost or low-cost meals provided through community meals or congregate
dining play an important role in decreasing food insecurity. Research in
Minnesota found that seniors experiencing food insecurity valued
congregate dining for providing affordable meals and a space for social
gathering (Oemichen and Smith, 2016). Furthermore, because seniors paid
for the meal, there was no stigma attached to the use of these services
compared to traditional food-purchasing assistance such as the
Supplemental Nutrition Assistance Program.

While previous research has examined the characteristics of individuals
and households accessing emergency and community food services, the
locational or transportation accessibility aspect of food bank access is
not well understood. A wealth of literature examining the food desert
concept suggests that, in addition to socio-economic and demographic
factors, location and transportation networks play a key role in a
household's accessibility to healthy foods (Paez et al., 2010; Vanderlee
and L'Abbé, 2017; Widener, 2018). For food banks specifically, previous
qualitative research in Ontario by Smith-Carrier et al. (2017) has noted
that ``transportation can be challenging, particularly if the food bank
is situated in a remote location'' (p.~32). Particularly, it appears
that participants experience challenges with the ``inordinate amount of
time necessary to obtain food, and difficulties associated with
transportation'' (p.~39). Users of food banks, according to this
research, rely on a variety of modes of transportation to access
services. Consequently, the location of facilities matters; in the words
of an interviewee: ``I wish it {[}the food bank{]} was a little more
centrally located. Because if I didn't have a bike I'd have to walk it
all the way out there and back. I wonder about people who don't''
(p.~39).

@@

More formally, the first step of this method is as follows:

\[
L_j = \frac{S_j}{\sum_{i=1}^nP_iw_{ij}}
\]

@@

The second step in this process is then summing the level of service
that each population unit can reach, according to the distance-decay
weight:

\[
A_i = \sum_{j=1}^JL_jw_{ji}
\]

@@

Floating catchment area methods are prone to overestimation of the
population and the level of service due to multiple-counting. The
population at \(P_i\) is allocated to \emph{every} service point \(j\)
for which \(w_{ij}>0\). Similarly, the level of service at \(LOS_j\) is
allocated to \emph{every} population point for which \(w_{ji}>0\). This
inflation effect has been known for several years, and several
modifications have been proposed to mitigate it (Delamater, 2013; e.g.,
Wan et al., 2012). A definitive solution to this issue was presented by
Paez et al. (2019). In order to avoid the multiple-counting in the
summations, the population and the level of service need to be allocated
\emph{proportionally}. This is achieved by standardizing the weights as
follows:

\[
w_{ij}^i = \frac{w_{ij}}{\sum_{i=1}^nw_{ij}}
\]

\noindent and:

\[
w_{ij}^j = \frac{w_{ij}}{\sum_{j=1}^Jw_{ij}}
\]

The standardized weights satisfy the following conditions:

\[
\sum_{i=1}^nw_{ij}^i=1
\]

\noindent and:

\[
\sum_{j=1}^Jw_{ij}^j=1
\]

Since the population is allocated proportionally, its value is
preserved:

\[
\sum_{i=1}^nP_iw_{ij}^i=P_i
\]

\noindent as is the level of service:

\[
\sum_{j=1}^JL_jw_{ij}^j=L_j
\]

@@

With a population of around 540,000, the City of Hamilton is the fourth
largest city in Ontario. It has historically been home to major
manufacturing industries but de-industrialization that has occurred over
the past several decades has led Hamilton to become one of the most
highly divided cities in Ontario, with a significant proportion of its
residents living at or below Canada's poverty level (DeLuca et al.,
2012; Jakar and Dunn, 2019; Latham and Moffat, 2007). The Hamilton
Community Foundation (HCF, 2018) reported that based on the Low-Income
Cut-Off, Hamilton recorded a poverty rate of 16.7\% in 2016, which was
well above the average rate of Ontario (13.7\%) and the average national
rate (12.8\%). According to data from Hamilton Food Share (HFS, 2019),
approximately 23,000 individuals accessed food banks in the city in
March of 2019. Within this total is 9,125 visits by children (minors up
to 18 years old), up from 8,278 the year before. Feed Ontario, the
province's largest collective of hunger-relief organizations, found that
on a per-capita basis, the level of need in the inner core of central
Hamilton was second highest in Ontario (FO, 2019).

@@

Data have been prepared for sharing in the form of an open data product
(see Arribas-Bel et al., 2021) available in the Supplemental Materials.
The contents of the data package are described next.

@@

Once we obtained travel time tables with population (number of
households) and proportion of trips by mode, we calculated the expected
travel time \(ett\) from each parcel \(i\) to a food bank or food
service location \(j\) as follows:

\[
ett_{ij} = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
\]

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Changes in accessibility to food banks and food services during
COVID-19 and implications for low income populations in Hamilton,
Ontario}
    \author[Some University]{Author One}
   \ead{author1@example.com} 
    \author[Another University]{Author Two\corref{1}}
   \ead{author2@example.com} 
    \author[Some University]{Author Three}
   \ead{author3@example.com} 
    \author[Some University]{Author Four}
   \ead{author4@example.com} 
      \address[Some University]{Department, Street, City, State, Zip}
    \address[Another University]{Department, Street, City, State, Zip}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  In this paper we analyze the changes in accessibility to food banks
  and related services before and during the COVID-19 pandemic in the
  City of Hamilton, Ontario. Food banks and services are the last line
  of support for households facing food insecurity; as such, their
  relevance cannot be ignored in the midst of the economic upheaval
  caused by the pandemic. Our analysis is based on the application of
  balanced floating catchment areas, and concentrates on households with
  lower incomes (\textless CAD40,000, approximately the Low Income
  Cutoff Value for a city of Hamilton's size). We find that
  accessibility was low to begin with in suburban and exurban parts of
  the city; furthermore, about 14\% of locations originally available in
  Hamilton closed during the pandemic, further reducing accessibility.
  The impact of closures on the level of service of the remaining
  facilities, and on accessibility, was disproportionate, with
  system-wide losses exceeding 40\%. Those losses were geographically
  and demographically uneven. While every part of the city faced a
  reduction in accessibility, inner suburbs fared worse in terms of loss
  of accessibility. As well, children (age \(\le 18\)) appear to have
  been impacted the most.
  \end{abstract}

@@

The emergence of COVID-19, the worst public health crisis since the 1918
flu pandemic, has exposed important social and economic fault lines, and
pre-existing patterns of inequality appear to have been exacerbated.
Along several other dimensions of stress (e.g., accessibility to health
care facilities, Ghorbanzadeh et al., 2021; Pereira et al., 2021a), this
seems to be the case for food insecurity as well (Laborde et al., 2020).
In the US, for example, it has been estimated that there was an increase
of more than 30\% in household food insecurity, and more than one third
of households were discovered to be newly food insecure - meaning they
did not experience food insecurity before the pandemic (Niles et al.,
2020). In Canada, Men and Tarasuk (2021) report that about 25\% of
individuals who experienced job insecurity (a relatively common
occurrence during the pandemic), also experienced food insecurity.
Similarly, according to Statistics Canada (2020a), in the early stages
of the pandemic almost 15\% of individuals reported living in a
household that faced food insecurity; the risk of food insecurity was
substantially higher for households with children. The difference
between households with and without children was significant, and 11.7\%
of households with children indicated that ``food didn't last and
{[}there was{]} no money to get more'' sometimes or often, compared to
7.3\% of households without children; likewise, 13\% of households with
children indicated that they ``{[}c{]}ouldn't afford balanced meals''
sometimes or often, compared to 8.8\% of households without children.

The impacts of food insecurity during the pandemic are alarming, since
diet-related diseases, such as obesity, heart-disease, and diabetes,
were already critical public health concerns in Canada prior to COVID-19
(Boucher et al., 2017). While food banks are not necessarily a stable
solution to food insecurity and in fact may encourage a retrenchment of
neoliberal policy (Wakefield et al., 2013), at least they can be argued
to provide a resource of last instance to households in precarious
situations (Bazerghi et al., 2016). As a mid-size city grappling with
deindustrialization, Hamilton exhibits high rates of poverty and food
bank use. As recently as 2019, the Hamilton Hunger Report\footnote{https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf}
noted that food banks in the city recorded the highest number of
visitors in the past 29 years; a rate of increase greater than
population growth. Most troubling, approximately 40\% of all visitors
were children.

It is known that urban food environments, within which people make their
daily food choices, are essential in influencing eating behaviours and
health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to food banks. Previous work
has explored differences in \emph{accessing} food banks, such as how
some households utilize food banks over short periods of time while
others regularly utilize food banks as longer-term resource (e.g., Enns
et al., 2020). In addition, transportation and locational considerations
have been raised as key issues in food bank accessibility in previous
qualitative research (Smith-Carrier et al., 2017). However, we are not
aware of any research that has focused on estimating or capturing this
geographic component of accessibility.

The study of place-based geographic accessibility is concerned with
capturing the potential to reach destinations of value using the
transportation network (Páez et al., 2012). Indeed, the Government of
Canada's recent Food Policy has made ``access'' to healthy food a
priority for Canadian communities\footnote{https://www.agr.gc.ca/eng/about-our-department/key-departmental-initiatives/food-policy/the-food-policy-for-canada/?id=1597863791042}
and previous survey research has suggested that such accessibility plays
a key role in user satisfaction with food bank service delivery (Holmes
et al., 2018). However, as with research into the prevalence of food
deserts, accessibility to food banks is unlikely to be evenly
distributed, and variation throughout a city can be expected due to
transportation network characteristics, and the spatial distribution of
food bank locations and the population they are meant to serve.
Furthermore, policy responses to the COVID-19 pandemic likely have added
to the distress of vulnerable households. Non-pharmaceutical
interventions during the pandemic involving restrictions in mobility
have increased the friction of travel, in particular by transit on which
low income populations are more reliant (e.g., DeWeese et al., 2020). At
the same time, the pandemic has created additional stress for the
operators of food banks through disruptions in the supply chain (e.g.,
McKay et al., 2021) as well as concerns surrounding the delivery of
service in safe conditions and possible cancellation of food service
programs.

For this study, we aim to look at how the landscape of food banks and
related services (e.g.~low-cost or free meal service providers)
available in Hamilton, Ontario, changed during the pandemic. Did the
number of open food bank services diminish? If so, what was the
accessibility to food banks before the pandemic from the perspective of
low income households, and how it changed during the pandemic? And
finally, who are most likely to have been impacted by changes in the
accessibility landscape? This paper first looks at the distribution of
food banks and related services before and during the pandemic. Then, we
use the balanced floating catchment area approach of Paez et al. (2019)
to investigate the accessibility situation. For this, we adopt a fully
disaggregated approach based on parcel-level data. Socio-economic and
demographic data are drawn from the latest Census of Canada (2016),
whereas travel information is from the most recent regional travel
survey from 2016. This paper follows reproducible research
recommendations (see Brunsdon and Comber, 2020), and the research was
conducted using open source tools for transportation analysis (Lovelace,
2021). The code and data necessary to reproduce the analysis are
available in a public repository\footnote{https://drive.google.com/drive/folders/1-l8hO1pMIqaBqf57j\_M\_BMrXhXFvrV2c?usp=sharing}.

\hypertarget{food-insecurity-and-food-bank-use-in-canada}{%
\section{Food Insecurity and Food Bank Use in
Canada}\label{food-insecurity-and-food-bank-use-in-canada}}

Food insecurity is the inability to acquire and consume an adequate
amount or good quality food, leading to inadequate nutrient intake (Enns
et al., 2020; Kirkpatrick and Tarasuk, 2008; Tarasuk and Vogt, 2009).
This nutrient deficiency is associated with major population health
concerns, particularly among Canadians at socio-economic disadvantage
(Bazerghi et al., 2016). Official government surveys such as the
Household Food Security Survey Module (HFSSM), the Canadian Community
Health Surveys (CCHS), the Longitudinal and International Study of
Adults (LISA), and official classifications determined by Health Canada
in relation to socio-demographic variables offer some insight into food
insecurity in Canada (Gundersen et al., 2018; Kirkpatrick and Tarasuk,
2008; Tarasuk and Vogt, 2009). Nationally, it has been found that food
insecurity impacts approximately 12.3\% of Canadian households (Tarasuk
et al., 2014)

Food banks - sometimes also referred to as `food pantries' and `food
shelves' - originated as a community response to aid those with
inadequate food by voluntarily offering them meals and ingredients
(Loopstra and Tarasuk, 2012; Riches, 2002). The scope and objectives of
food banks can vary by region and by country, and these organizations
can include not only prepared meals and aliments, but also shared spaces
to connect in community gardens and community kitchens (Wakefield et
al., 2013). Although in their origin food banks were meant to be provide
a temporary solution to accommodate those in hunger due to job
retrenchments and economic downfalls since the 1980s, over time they
have evolved into a community practice to secure emergency food supplies
for those in need (Loopstra and Tarasuk, 2012; Wakefield et al., 2013).

In Canada, the number of food banks has steadily increased in the past
few decades (Wakefield et al., 2013). The largest database of food banks
and their use comes from the non-profit association Food Banks Canada
(FBC), which conducts an annual assessment through its affiliated
members. FBC's 2018 Hunger Count report\footnote{https://foodbankscanada.ca/getmedia/241fb659-05f5-44a2-9cef-56f5f51db523/HungerCount-2018\_FINAL\_EN.pdf.aspx?ext=.pdf}
(the most recent available) listed 1,830 member food banks across the
country, and found that Canadians visited food banks 1.1 million times
in March of 2018. Of those accessing food banks, certain population
characteristics tend to be over-represented compared to national totals
from the 2016 Canadian Census of Population. According to FBC's 2018
data, single-adult households represent 45\% of those utilizing food
banks despite making up 28\% of Canada's population, 19\% are
single-parent households (compared to 10\% nationally), and 35\% of
those accessing food bank services are children aged 0-18 even though
their share of Canada's national population is approximately 20\%. In
addition, 59\% of households accessing food banks list social or
disability assistance as their primary source of income. Similarly,
using data from the 2011-2012 CCHS, Tarasuk et al. (2019) found higher
odds of food insecurity amongst households relying on social assistance,
those without a university degree or with children under the age of 18,
and individuals that lived alone, renters, and those identifying as
Aboriginal. While surveys revealed that only 20 to 30 percent of those
experiencing food insecurity were found to frequent food banks in Canada
(Tarasuk et al., 2014), research from Ottawa (Enns et al., 2020) and
Vancouver (Black and Seto, 2020) suggests that long-term users tend to
be older, have health or mobility challenges, live in large households,
and are less likely to have employment income.

Economic stress caused by the COVID-19 pandemic, rising unemployment
rates, and changes in poverty levels have disrupted the food environment
and led to higher rates of food insecurity (Niles et al., 2020).
Pandemic-related food security studies in the US have found a
substantial increase in households experiencing food insecurity for the
first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects,
increases in food insecurity rates due to the onset of COVID-19, signal
a change in the food environment with potential damages to population
health during the course of the pandemic and beyond (Niles et al.,
2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from FBC showed that 52\% of member food banks reported an increase in
usage in March of 2020 when initial lockdown restrictions were put in
place across much of the country. The pandemic also created significant
staffing issues with 42\% of food banks reporting a reduction in
volunteers. However, 53\% of food banks later reported a decrease in use
into the summer of 2020 which FBC members attributed to emergency
financial support programs from the federal government. Nevertheless,
some of these benefit programs were temporary. Although more recent
statistics on food bank use in Hamilton in 2020 and 2021 are not yet
available, data from the Daily Bread Food Bank in neighbouring Toronto
for August 2020 shows visits climbing 51\% year-over-year in that
city\footnote{https://www.dailybread.ca/wp-content/uploads/2020/11/Whos-Hungry-Report-2020.pdf},
which suggests that many households in Hamilton are likely to turn to
food bank services to meet their needs.

In terms of geography, previous research conducted at the provincial
scale using data from the 2011-2012 CCHS found that the prevalence of
food insecurity ranged across the country from 11.8\% of households in
Ontario to 41\% of households in Nunavut (Tarasuk et al., 2019).
However, although previous research has examined the characteristics of
individuals and households accessing food banks, the locational or
transportation accessibility aspect of food bank access is not well
understood. A wealth of literature examining the food desert concept
suggests that, in addition to socio-economic and demographic factors,
location and transportation networks play a key role in a household's
accessibility to healthy foods (Paez et al., 2010; Vanderlee and L'Abbé,
2017; Widener, 2018). For food banks specifically, previous qualitative
research in Ontario by Smith-Carrier et al. (2017) has noted that
``transportation can be challenging, particularly if the food bank is
situated in a remote location'' (p.~32). Particularly, it appears that
participants experience challenges with the ``inordinate amount of time
necessary to obtain food, and difficulties associated with
transportation'' (p.~39). Users of food banks, according to this
research, rely on a variety of modes of transportation to access
services. Consequently, the location of facilities matters; in the words
of an interviewee: ``I wish it {[}the food bank{]} was a little more
centrally located. Because if I didn't have a bike I'd have to walk it
all the way out there and back. I wonder about people who don't''
(p.~39).

@@

More formally, the first step of this method is as follows: \[
L_j = \frac{S_j}{\sum_{i=1}^nP_iw_{ij}}
\]

@@

The second step in this process is then summing the level of service
that each population unit can reach, according to the distance-decay
weight: \[
A_i = \sum_{j=1}^JL_jw_{ji}
\]

@@

Floating catchment area methods are prone to overestimation of the
population and the level of service due to multiple-counting. The
population at \(P_i\) is allocated to \emph{every} service point \(j\)
for which \(w_{ij}>0\). Similarly, the level of service at \(LOS_j\) is
allocated to \emph{every} population point for which \(w_{ji}>0\). This
inflation effect has been known for several years, and several
modifications have been proposed to mitigate it (Delamater, 2013; e.g.,
Wan et al., 2012). A definitive solution to this issue was presented by
Paez et al. (2019). In order to avoid the multiple-counting in the
summations, the population and the level of service need to be allocated
\emph{proportionally}. This is achieved by standardizing the weights as
follows: \[
w_{ij}^\text{st} = \frac{w_{ij}}{\sum_{i=1}^nw_{ij}}
\] \noindent and: \[
w_{ji}^\text{st} = \frac{w_{ji}}{\sum_{j=1}^Jw_{ji}}
\]

The standardized weights satisfy the following conditions: \[
\sum_{i=1}^nw_{ij}^\text{st}=1
\]

\noindent and: \[
\sum_{j=1}^Jw_{ji}^\text{st}=1
\]

Since the population is allocated proportionally, its value is
preserved: \[
\sum_{i=1}^nP_iw_{ij}^\text{st}=P_i
\]

\noindent as is the level of service: \[
\sum_{j=1}^JL_jw_{ji}^\text{st}=L_j
\]

@@

With a population of around 540,000, the City of Hamilton is the fourth
largest city in Ontario. It has historically been home to major
manufacturing industries but de-industrialization that has occurred over
the past several decades has led Hamilton to become one of the most
highly divided cities in Ontario, with a significant proportion of its
residents living at or below Canada's poverty level (DeLuca et al.,
2012; Jakar and Dunn, 2019; Latham and Moffat, 2007). In 2016, the
Hamilton Community Foundation reported that based on the Low-Income
Cut-Off, Hamilton recorded a poverty rate of 16.7\%, which was well
above the average rate of Ontario (13.7\%) and the average national rate
(12.8\%).\footnote{https://www.hamiltoncommunityfoundation.ca/vital-signs/low-income-2018/}.
According to data from Hamilton Food Share\footnote{https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf},
approximately 23,000 individuals accessed food banks in the city in
March of 2019. Within this total is 9,125 visits by children (minors up
to 18 years old), up from 8,278 the year before. Feed Ontario, the
province's largest collective of hunger-relief organizations, found that
on a per-capita basis, the level of need in the inner core of central
Hamilton was second highest in Ontario\footnote{https://feedontario.ca/research/report-hunger-map/}.

@@

Data have been prepared for sharing as a data package\footnote{https://drive.google.com/drive/folders/1-l8hO1pMIqaBqf57j\_M\_BMrXhXFvrV2c?usp=sharing}.
The contents of the data package are described next.

@@

Once we obtained travel time tables with population (number of
households) and proportion of trips by mode, we calculated the expected
travel time \(ett\) from each parcel \(i\) to a food bank or food
service location \(j\) as follows: \[
ett_{ij} = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
\]
